% Beamer Presentation: Lisbon Conference Talk
\documentclass[aspectratio=169]{beamer}

% Theme and colors - customize as needed
\usetheme{Madrid}
\usecolortheme{default}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{tikz}

% Bibliography
\usepackage[backend=biber,style=authoryear]{biblatex}
\addbibresource{references.bib}

% Metadata
\title{From Plausibility Machines to Narrative Minds}
\subtitle{Vision, Attention, and the Architecture of Artificial Cognition}
\author{Brook Miller}
\institute{Center for Humanities and Digital Research, University of Central Florida}
\date{\today}

\begin{document}

% Title slide
\begin{frame}
  \titlepage
\end{frame}

% Outline
\begin{frame}{Outline}
  \tableofcontents
\end{frame}

%=============================================================================
\section{Plausibility Machines and the Bootstrapping Problem}
%=============================================================================

\begin{frame}{Word Search Junkies}
  \begin{itemize}
    \item Since ChatGPT's release (Dec.\ 2022), a wave of claims that AGI is imminent
    \item<2-> Pushback: Chiang and Tufekci --- LLMs as \alert{plausibility machines} \parencite{chiang_plausibility,tufekci2026hype}, producing text that reads as coherent without underlying understanding
  \end{itemize}

  \pause
  \vspace{1em}
  \begin{block}{Long's reframing \parencite{long2026experience}}
    \textbf{Sneerer:} ``just predicting the next token'' $\Rightarrow$ therefore not conscious \\
    \textbf{Defender:} humans are unreliable and contradictory too $\Rightarrow$ so what's the difference?
  \end{block}

  \pause
  \vspace{1em}
  \begin{alertblock}{The more interesting question}
    Not \emph{whether} LLMs and humans differ, but precisely \emph{how} --- and what that implies.
  \end{alertblock}
\end{frame}

\begin{frame}{The Bootstrapping Problem}
  \begin{block}{Long \parencite{long2026experience}}
    LLMs ``learned language first and would bootstrap any potential experiences from that, whereas humans can have experiences independently of using language to describe them.''
  \end{block}

  \pause
  \vspace{1em}
  \begin{itemize}
    \item<2-> It may make sense to ask what it is like to be a bat; asking the same of an LLM is dubious
    \item<3-> \textbf{What an LLM is:} procedural circuits --- comparable to functions in a CS sense
    \item<4-> \textbf{What is (partly) missing:} valence and salience
  \end{itemize}

  \pause
  \vspace{1em}
  \begin{alertblock}{Claim}
    If we want AI with a closer semblance to human cognition, a language-first architecture may be the wrong place to start. Look elsewhere.
  \end{alertblock}
\end{frame}

%=============================================================================
\section{An Analogy: Human and Computer Vision}
%=============================================================================

\begin{frame}{An Analogy: Human and Computer Vision}
  \begin{itemize}
    \item<1-> \textbf{Edge detection} --- low-level feature extraction
    \item<2-> \textbf{Abstraction} --- features aggregate into intermediate representations
    \item<3-> \textbf{Object recognition} --- abstraction resolves into a labeled category
  \end{itemize}

  \pause
  \vspace{1em}
  \begin{block}{Why look here}
    Vision research offers embodied, non-linguistic experience-formation to study --- a candidate escape from the bootstrapping problem. Raw input $\rightarrow$ abstraction $\rightarrow$ recognition is a useful template for thinking about cognitive architecture generally. Where does narrative fit in this pipeline?
  \end{block}
\end{frame}

%=============================================================================
\section{Event Segmentation: A Basis for Experience}
%=============================================================================

\begin{frame}{Oh No, Waymo}
  \begin{itemize}
    \item Waymos endlessly circling roundabouts and neighborhoods --- fair grounds for skepticism
    \item<2-> But moving, embodied AI does not \emph{ipso facto} share the LLM's bootstrapping problem
    \item<3-> Regulating a body moving through space is an ongoing, dynamic problem: recognizing other agents and dynamic forces in the world
  \end{itemize}

  \pause
  \vspace{1em}
  \begin{block}{Slow feature analysis \parencite{song2022slow}}
    A baseline for experiential processing --- it builds temporal parsing into experiencing.
  \end{block}
\end{frame}

\begin{frame}{Event Segmentation}
  \begin{block}{What an event is, minimally}
    ``a segment of time at a given location that is perceived by an observer to have a beginning and an end'' \parencite[p.~3]{zackstversky2001event}
  \end{block}

  \pause
  \vspace{1em}
  \begin{block}{Event Segmentation Theory (EST)}
    ``One central proposal of the EST model is that event segmentation is a spontaneous outcome of ongoing perception and arises at points where perceptual predictions begin to fail or change\ldots The period of epoch between two such consecutive changes current mental models is an event.'' \parencite[p.~591]{shipleyzacks2008understanding}
  \end{block}
\end{frame}

\begin{frame}{Prediction, Revisited}
  \begin{itemize}
    \item Sneerer/defender critique centers on LLMs' prediction as reductive: next token, given previous tokens
    \item<2-> But perceptual, event-driving prediction is different in kind:
  \end{itemize}

  \pause
  \vspace{1em}
  \begin{block}{}
    ``As an event proceeds, predictions are relatively accurate. However, as it ends, prediction error increases, causing a reset of the event model. When such a reset occurs, an event boundary is set.'' \parencite[p.~135]{miller2023eventsweaknarrativity}
  \end{block}

  \pause
  \vspace{1em}
  \begin{alertblock}{This requires online processing of ongoing experience}
    Two research threads follow: \textbf{slow feature analysis} and \textbf{uncertainty processing}.
  \end{alertblock}
\end{frame}

%=============================================================================
\section{``Attention Is All You Need''}
%=============================================================================

\begin{frame}{``Attention Is All You Need'' \parencite{vaswani2017attention}}
  \begin{itemize}
    \item What attention consists of, for an LLM
    \item Stacked transformers and attention
    \item Loss minimization and static associative valuation
  \end{itemize}
\end{frame}

\begin{frame}{Attention, Contrasted: Weights vs.\ Awareness}
  \begin{itemize}
    \item<1-> Transformer attention: a computed weighting over tokens in a context window --- not modeled as an object of the system's own awareness
    \item<2-> Webb and Graziano's \textbf{Attention Schema Theory} \parencite{webbgraziano2015attention}: awareness is \emph{an internal model of attention}
  \end{itemize}

  \pause
  \vspace{1em}
  \begin{block}{``A creature with an attention schema...''}
    ``...should be a creature that concludes it is aware'' \parencite{webbgraziano2015attention}
  \end{block}

  \pause
  \vspace{1em}
  \begin{alertblock}{Open possibility}
    Could an explicit, internal model \emph{of the system's own attention} --- not just attention weights themselves --- point toward AI architectures with a closer semblance of awareness? (cf.\ Section ``What Architecture Is Closer?'')
  \end{alertblock}
\end{frame}

%=============================================================================
\section{Coglinks and Uncertainty}
%=============================================================================

\begin{frame}{Coglinks and Uncertainty}
  \begin{block}{Second research thread, promised above}
    Uncertainty processing in hierarchical decision-making \parencite{wang2025neural}
  \end{block}

  \pause
  \vspace{1em}
  \begin{itemize}
    \item Circuits imitated
    \item Predefined decision choices
  \end{itemize}

  % NOTE (to develop): CogLink advances neural modeling of hierarchical, model-based/model-free
  % decision-making, but does not address *ill-defined problems* -- where no predefined script or
  % solution path applies. My Ch. 8 material frames this as a gap "vexing for schema theory"
  % generally; since CogLink is in effect a neural schema/model-coordination account, the same gap
  % plausibly applies to it. Weak-narrative dynamics (gist extraction, event simulation) are a
  % candidate mechanism for what keeps functioning when the model itself has no applicable script.
  % See quotes.md, Miller Ch. 8 section, (b) intervention point on ill-defined problems.
\end{frame}

%=============================================================================
\section{Time, Prediction, and Memory}
%=============================================================================

\begin{frame}{Time, Prediction, and Memory}
  \begin{itemize}
    \item Predictive processing vs.\ uncertainty
    \item Episodic and semantic memory
  \end{itemize}
\end{frame}

%=============================================================================
\section{Phenomenological Experience and Emotion}
%=============================================================================

\begin{frame}{Phenomenological Experience and Emotion}
  \begin{block}{Open question}
    What would have to be true of a system for phenomenological experience and emotion --- not merely their outward signatures --- to be present?
  \end{block}

  % Add discussion points
\end{frame}

%=============================================================================
\section{Narrative in Cognition}
%=============================================================================

\begin{frame}{Narrative in Cognition \parencite{miller2024narrativity}}
  \begin{itemize}
    \item<1-> As \textbf{active inference}
    \item<2-> As an \textbf{abstraction scaffold}
  \end{itemize}
\end{frame}

%=============================================================================
\section{What Architecture Is Closer?}
%=============================================================================

\begin{frame}{What Architecture Is Closer?}
  \begin{itemize}
    \item Self-salience models
    \item Less terrifying possibilities that simulate emotion
  \end{itemize}

  % NOTE: "Self-salience models" connects back to Attention Schema Theory (Webb & Graziano),
  % introduced in the "Attention Is All You Need" section as a contrast to transformer attention --
  % an explicit internal model of one's own attention is itself a self-salience mechanism.
\end{frame}

%=============================================================================
\section{What Training Is Required?}
%=============================================================================

\begin{frame}{What Training Is Required?}
  \begin{itemize}
    \item Direction 1: [To be developed]
    \item Direction 2: [To be developed]
  \end{itemize}
\end{frame}

%=============================================================================
\section{How Might a Narrative-Processing AI Differ?}
%=============================================================================

\begin{frame}{How Might a Narrative-Processing AI Differ?}
  \begin{enumerate}
    \item<1-> Not procedural?
    \item<2-> Predictive processing (with respect to dynamic inputs)?
    \item<3-> Nested experiential hierarchies --- including reflection on what experiences reveal about self-function?
  \end{enumerate}

  \pause
  \vspace{1em}
  \begin{block}{Thus}
    \alert{A closer semblance of self.}
  \end{block}
\end{frame}

%=============================================================================
\section{Notes / Parking Lot}
%=============================================================================

\begin{frame}[allowframebreaks]{Notes / Parking Lot}
  % Scratch space for ideas, anecdotes, and examples to develop or place elsewhere
  \footnotesize
  \begin{itemize}
    % "Oh no Waymo" promoted to a full frame in Section "Event Segmentation: A Basis for Experience".
    % Retained here: the valuative/attentional follow-on not yet worked into that frame.
    \item \textbf{Waymo, continued} --- % TODO: develop
      \\ \emph{valuative processing and attention, beyond slow feature analysis}

      \vspace{0.5em}
      Slow feature analysis combines with valuative processing (salience and valence) that is itself dynamic and interactive. In terms of attention, a stabilization in valuation can be a catalyst of shifting attention, which marks temporal experience with subjective meaning.

      \vspace{0.5em}
      \emph{Open questions:} Can an experiencing AI have a semblance of these processes? How might it relate to the question of whether the AI has subjective experience?

    \item \textbf{Hebbian learning vs.\ gradient descent} --- % TODO: develop
      \\ \emph{Hebbian as local; gradient descent as global}
  \end{itemize}
\end{frame}

%=============================================================================
\section{Conclusion}
%=============================================================================

\begin{frame}{Thank You}
  \centering
  \textbf{Thank you!}

  \vspace{0.5em}

  \textit{Questions?}
\end{frame}

%=============================================================================
% References
%=============================================================================

\begin{frame}[allowframebreaks]{References}
  \printbibliography
\end{frame}

\end{document}
